Practical Dependently-Typed Programming for the Web

### 2013-08-21 paper/main.tex
Paper revisions

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
\usetikzlibrary{arrows,automata}
%\WarningsOff

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}

@@

\begin{abstract}
Dependently-typed languages allow for much expressive types to be used within
programs in turn allowing easier reasoning about the operation of such
programs. Stronger specifications do however bring with them the disadvantage
that it becomes increasingly difficult to write programs that are accepted by
the type checker and additional proofs may have to be specified by a user.

@@

In this paper, we apply this technique to web programming, and introduce an
EDSL, to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

@@

\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. At
this point, a developer may forget to execute the statement before beginning to
fetch results or may have forgotten to bind the parameters to the statement:

@@

This complexity can be ameliorated through the use of \textit{embedded
domain-specific languages} (EDSLs), which aim to abstract away the
complexity of the underlying type system. EDSLs allow domain experts to
focus on a problem domain, rather than the specific details of their
solution.

@@

\subsection{Contributions}
The primary contribution of this paper is the application of 
dependent types may to provide strong static guarantees
about the correctness and security of web applications, whilst minimising
additional development complexity. In particular, we present:

\begin{itemize}
\item A form-handling mechanism, which preserves type information and
manages user input, therefore
increasing applications' resilience to attacks such as SQL injection.

\item Representations of CGI, Databases and sessions as
\textit{algebraic effects}, allowing programs to be accepted only when they
follow clearly defined resource usage protocols.

\item A practical demonstration of the framework, a message board application.

@@

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; implement
programs which, by using this effet, implicitly carry a proof that the resource
protocol has been correctly followed.

@@

CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client. Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.

@@

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure~\ref{fig:cgistates} shows the states through which the CGI program
progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
as a dependent algebraic effect. Each operation performed in an effectful
program requires the resource to be of a certain type, and the completion of
the operation may alter the type or value of the resource.

Upon creation, the CGI application is uninitialised, meaning that the
environment variables and request have not been retrieved. The only operation
that can be performed in this state is initialisation: by calling
\texttt{initialise}, a CGIInfo record is populated, and the state transitions
to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
the initialised state.

@@

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request. Finally, the headers and content are written which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. This is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity.

@@

\begin{Verbatim}
'SELECT `name`, `address` FROM `staff` LIMIT ' + 
       page + ', 5';
\end{Verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be
modified by an attacker to include a malicious string which would alter the
semantics of the query, allowing an attacker to execute a blind SQL injection
attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
clauses here

@@

\subsection{Form Construction}
\label{formcons}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
\begin{Verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{Verbatim}
Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
\begin{Verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{Verbatim}
%
In order to specify a form, we once again use \texttt{Effects}. By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: the specification of an
incompatible handler will result in a compile-time type error.

@@

We make use of the resource associated with the effect, \texttt{FormRes}, to
construct the form. The resource allows us to record the types associated with
each form element and the HTML required to display the form. he constructor,
\texttt{FR}, requires the number of elements in the form in order to allow for
the naming of new elements, the list of element types, the list of effects
supported by the handler function, and the currently generated HTML for the
form.

@@

To specify a form instance, we define a function of type \texttt{UserForm}:
\begin{Verbatim}
  UserForm : Type
  UserForm = EffM m [FORM (FormRes []) 
                          (FormRes [])] String
\end{Verbatim}
All forms are required to include a submit button, as mandated by the
requirement that the input and output resource contains an empty list of types;
this requirement is fulfilled as per the output resource type of the
\texttt{AddSubmit} operation. As the creation of a form is a pure function
which does not include side effects, we do not restrict the handler to IO as
with previously-discussed effects, instead denoting the fact that it may be run
in any handler with the implicit variable \texttt{m}.

@@

Before associating a handler function with the form, we must specify the effects available to the handler. This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource. By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form. 
\begin{Verbatim}
  useEffects : (effs : List WebEffect) ->
               EffM m [FORM (FormRes G E)] 
                      [FORM (FormRes G effs)] ()
  useEffects effs = (UseEffects effs)
\end{Verbatim}
Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}. This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:
\begin{Verbatim}
  FormHandler : List EFFECT -> Type
  FormHandler effs = Eff IO effs ()
\end{Verbatim}
In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
\begin{Verbatim}
  addSubmit : 
    (f :  mkHandlerFn ((reverse G), E)) ->
    (fns : HandlerList) ->
    {default tactics 
      { applyTactic findFn 100; solve; }
      prf : FnElem f fns} ->
    EffM m [FORM (FormRes G E)]
           [FORM (FormRes [] [])] 
           String
  addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf          
\end{Verbatim}

@@

Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page. Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}. 
\subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations. 
\begin{Verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{Verbatim}
We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled. If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
\begin{Verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{Verbatim}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
\begin{Verbatim}
newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers
\end{Verbatim}
The form consists of two text boxes: one for the title of the thread, and one for the content of the first post. Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value. The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
\begin{Verbatim}
handlers : HandlerList
handlers = [
 (handler args=[FormString, FormString], 
          effects=[CgiEffect, SessionEffect, SqliteEffect], 
          fn=handleRegisterForm, 
          name="handleRegisterForm"),

@@

  pure ()

@@

    Left err => pure False

@@

traverseThreads [] = pure ()

@@

        traverseRow _ = pure ()

@@

Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
dependent types to enable the generation of provably-correct and secure queries
and DOM code, without requiring overly complex types or proofs. Through the use
of the Ur\/Web DSL, developers may generate well-typed code from records. By
statically verifying the \textit{code generator}, it is therefore possible to
assert that the generated code is secure. Such ideas are extremely interesting
for the implementation of object-relational mapping, which we hope to implement
in the near future.

@@

In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers. With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available. Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.

@@

Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation. Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures. If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error. This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step. Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness. Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.

>>>>

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
%\WarningsOff

@@

\conferenceinfo{IFL '13}{August 28--31, 2013, Nijmegen, The Netherlands} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

\title{Practical Dependently-Typed Programming for the Web}
\subtitle{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}

@@

\begin{abstract}
Dependently-typed languages allow very expressive types to be used during development, in turn facilitating easier reasoning about the operation of programs written in such languages. Stronger type specifications do however bring with them the disadvantage
that it becomes increasingly difficult to write programs that are accepted by
the type checker and additional proofs may have to be specified by a user.

@@

In this paper, we apply this technique to web programming, and introduce an
EDSL to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

@@

\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. 
Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.

@@

This complexity can be ameliorated through the use of \textit{embedded
domain-specific languages} (EDSLs), which aim to abstract away the
complexity of the underlying type system. EDSLs allow domain experts to
write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.

@@

\subsection{Contributions}
The primary contribution of this paper is the application of 
dependent types to provide strong static guarantees
about the correctness and security of web applications, whilst minimising
additional development complexity. In particular, we present:

\begin{itemize}
\item A form-handling mechanism, which preserves type information and
manages user input, therefore
increasing applications' resilience to attacks such as SQL injection and cross-site scripting.

\item Representations of CGI, Databases and sessions as
\textit{resource-dependent algebraic effects}, allowing programs to be accepted only when they
follow clearly defined resource usage protocols.

\item A message board application, demonstrating the usage of the framework.

@@

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; implement
programs which, by using this effect, implicitly carry a proof that the resource
protocol has been correctly followed.

@@

CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client. Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error being shown.

@@

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure~\ref{fig:cgistates} shows the states through which the CGI program
progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
as a resource-dependent algebraic effect. Each operation performed in an effectful
program requires the resource to be of a certain type, and the completion of
the operation may alter the type or value of the resource.

Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
that can be performed in this state is initialisation: by calling
\texttt{initialise}, a CGIInfo record is populated, and the state transitions
to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
the initialised state.

@@

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, and the headers are written.
Finally, the headers and content are written which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. This is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity).

@@

\begin{Verbatim}
'SELECT `name`, `address` FROM `staff` LIMIT ' + 
       page + ', 5';
\end{Verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be
modified by an attacker to include a malicious string which would alter the
semantics of the query, allowing an attacker to execute a blind SQL injection
attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
% clauses here

@@

\subsection{Form Construction}
\label{formcons}
Each form element is specified  to hold a particular type of data, which, assuming that the correct type of data is specified by the user, is passed directly to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
\begin{Verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{Verbatim}
Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
\begin{Verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{Verbatim}
%
In order to specify a form, we once again use \texttt{Effects}. By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: the specification of an
incompatible handler will result in a compile-time type error.

@@

We make use of the resource associated with the effect, \texttt{FormRes}, to
construct the form. The resource allows us to record the types associated with
each form element and the HTML required to display the form. The constructor,
\texttt{FR}, requires the number of elements in the form in order to allow for
the naming of new elements, the list of element types, the list of effects
supported by the handler function, and the currently generated HTML for the
form.

@@

To specify a form instance, we define a function of type \texttt{UserForm}:
\begin{Verbatim}
  UserForm : Type
  UserForm = EffM m [FORM (FormRes []) 
                          (FormRes [])] String
\end{Verbatim}
All forms are required to include a submit button, as mandated by the
requirement that the input and output resource contains an empty list of types;
this requirement is fulfilled as per the output resource type of the
\texttt{AddSubmit} operation. As the creation of a form is a return ()function
which does not include side effects, we do not restrict the handler to IO as
with previously-discussed effects, instead denoting the fact that it may be run
in any handler with the implicit variable \texttt{m}.

@@

Before associating a handler function with the form, we must specify the effects available to the handler. This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource. By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form. 
\begin{Verbatim}
  useEffects : (effs : List WebEffect) ->
               EffM m [FORM (FormRes G E)] 
                      [FORM (FormRes G effs)] ()
  useEffects effs = (UseEffects effs)
\end{Verbatim}
Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}. This is, however, not an inherent limitation as the \texttt{Effects} library permits introduction of additional effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:
\begin{Verbatim}
  FormHandler : List EFFECT -> Type
  FormHandler effs = Eff IO effs ()
\end{Verbatim}
In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
\begin{Verbatim}
  addSubmit : 
    (f :  mkHandlerFn ((reverse G), E)) ->
    (fns : HandlerList) ->
    {default tactics 
      { applyTactic findFn 100; solve; }
      prf : FnElem f fns} ->
    EffM m [FORM (FormRes G E)]
           [FORM (FormRes [] [])] 
           String
  addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf          
\end{Verbatim}

@@

Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page. Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}. 
\subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations. 
\begin{Verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          return ()
\end{Verbatim}
We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled. If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
\begin{Verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.return ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{Verbatim}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
\begin{Verbatim}
newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers
\end{Verbatim}
The form consists of two text boxes: one for the title of the thread, and one for the content of the first post. Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value. The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
\begin{Verbatim}
handlers : HandlerList
handlers = [
 (handler args=[FormString, FormString], 
          effects=[CgiEffect, SessionEffect, SqliteEffect], 
          fn=handleRegisterForm, 
          name="handleRegisterForm"),

@@

  return ()

@@

    Left err => return False

@@

traverseThreads [] = return ()

@@

        traverseRow _ = return ()

@@

Ur\/Web \cite{urweb} is a library built for the Ur language, which does not use dependent types but does have a more expressive type system, making extensive use of record types and type-level computation.
By using these concepts, Ur\/Web may generate provably correct and unexploitable DOM code and SQL queries from records, without requiring developers to supply proofs. 
In contrast to using runtime code generation, which is prone to obscure code generation errors, Ur\/Web makes use of its static type system to guarantee that metaprograms---in this case, generated SQL and DOM code---must be correct and secure.
Such ideas regarding the use of static checking of metaprogram generation will be extremely useful when considering an object-relational mapping system, which we hope to implement in the near future. It will also be interesting to see how such concepts may be applied with a yet more expressive type system involving full dependent types.
%Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
%dependent types to enable the generation of provably-correct and secure queries
%and DOM code, without requiring overly complex types or proofs. Through the use
%of the Ur\/Web DSL, developers may generate well-typed code from records. By
%statically verifying the \textit{code generator}, it is therefore possible to
%assert that the generated code is secure. Such ideas are extremely interesting
%for the implementation of object-relational mapping, which we hope to implement
%in the near future.

@@

In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers. With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available. Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error or memory access violation.

@@

Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used during compilation. Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures. If data structures within the program do not conform to a given database schema, for example, then the program will not type-check and as such will produce a compile-time error. 
This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step. Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness. Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
